Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-01 biology-data.tex
preferable citation sorting

PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocites{smithPCR}{garibyanPCR}.
This is known as \emph{amplifying} the DNA.
PCR has many uses within the biological sciences; Mark R.\ Hughes, deputy director of the Human Genome Project, has claimed it is the most important scientific technology of the last hundred years~\autocite{powledgePCR}.
It is central to the work in this thesis because it can detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.

@@

PCR consists of a series of cycles, each cycle containing a three-step process~\autocites{powledgePCR}{garibyanPCR}.
The first step of the PCR cycle is \emph{denaturation}.
In denaturation, the DNA is heated to separate the two strands of DNA.
The second step is \emph{annealing}.
In annealing, the temperature is lowered to allow primers to bind to the DNA.
Primers are short sequences of DNA that are complementary to the DNA sequence of interest.
By choosing the primers, the PCR process targets a portion of DNA starting and ending with a specified sequence.
The final step is \emph{extension}.
In extension, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
In the course of such a cycle, the amount of DNA doubles.
Therefore, the number of DNA copies increases geometrically with the number of cycles completed.

>>>>

PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocite{smithPCR,garibyanPCR}.
This is known as \emph{amplifying} the DNA.
PCR has many uses within the biological sciences; Mark R.\ Hughes, deputy director of the Human Genome Project, has claimed it is the most important scientific technology of the last hundred years~\autocite{powledgePCR}.
It is central to the work in this thesis because it can detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.

@@

PCR consists of a series of cycles, each cycle containing a three-step process~\autocite{powledgePCR,garibyanPCR}.
The first step of the PCR cycle is \emph{denaturation}.
In denaturation, the DNA is heated to separate the two strands of DNA.
The second step is \emph{annealing}.
In annealing, the temperature is lowered to allow primers to bind to the DNA.
Primers are short sequences of DNA that are complementary to the DNA sequence of interest.
By choosing the primers, the PCR process targets a portion of DNA starting and ending with a specified sequence.
The final step is \emph{extension}.
In extension, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
In the course of such a cycle, the amount of DNA doubles.
Therefore, the number of DNA copies increases geometrically with the number of cycles completed.

### 2024-02-01 cis-perfect-testing.tex
easy edits
preferable citation sorting

Few studies look at the combination of both double interval censoring and undetected (or truncated) events
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}

@@

Bayesian methods been used for nest survival before~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible here.
I develop a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

@@

In this section, I develop a model accounting for double interval censoring and undetected episodes.
This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
This framework is itself based long-standing methodology~\autocites{dempsterMaximum}{turnbullEmpirical}.
I augment the data set with the full number of episodes, including those that are undetected.

@@

\todo{Not sure if the followin paragraph is understandable}
In forming the probability $p_{ia}$, the data is used.
This is not quite correct because forming a likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

@@

Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
The discrete Beta process prior generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time~\autocites{ibrahimBayesian}{sunStatisticala}.
It is:
\begin{align}
  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
  \alpha_t &= k_t h_t + \alpha_0 \\
  \beta_t &= k_t (1 - h_t) + \beta_0
\end{align}
where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
An intuition for what this distribution represents can be gained from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).

@@

Various extensions could be worth exploring.
Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocites{hakkiOnset}{russellWithinhost}.
Covariates can be included with a variety of models, the simplest being a proportional hazards model.
A major challenge here is that the calculation of $p_t$ is greatly complicated, because it is no longer linear in the survival function.

>>>>

Few studies look at the combination of both double interval censoring and undetected (or truncated) events
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}.
The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}

@@

Bayesian methods been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling}, augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible here.
I develop a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

@@

In this section, I develop a model accounting for double interval censoring and undetected episodes.
This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
I augment the data set with the full number of episodes, including those that are undetected.

@@

In forming the probability $p_{ia}$, the data is used.
This is not quite correct because forming a likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

@@

Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
The discrete Beta process prior generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time~\autocite{ibrahimBayesian,sunStatisticala}.
It is:
\begin{align}
  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
  \alpha_t &= k_t h_t + \alpha_0 \\
  \beta_t &= k_t (1 - h_t) + \beta_0
\end{align}
where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
An intuition for what this distribution represents can be gained from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).

@@

Various extensions could be worth exploring.
Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocite{hakkiOnset,russellWithinhost}.
Covariates can be included with a variety of models, the simplest being a proportional hazards model.
A major challenge here is that the calculation of $p_t$ is greatly complicated, because it is no longer linear in the survival function.

### 2024-02-01 incidence-prevalence.tex
preferable citation sorting

The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocites{beckerDependent}{brookmeyerMethod}.

@@

% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
% A super-spreading event is an event where many more people than average are infected.
% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
% %The negative binomial tends to Poisson as $k\to\infty$.
% Generally, $k<1$ is considered as significant levels of overdispersion.
% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling}{andersonSpread}.
This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The system can now be written as follows:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocites{birrellBayesian}{jacksonEffects}.
Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12. A clear diagonal, showing associative mixing is visible, and children have more contacts. Parallel diagonals off the main diagonal show mixing between parents and their children. Figure adapted from \textcite{klepacContacts}, licensed under \href{https://creativecommons.org/licenses/by-nc-nd/4.0/}{CC BY-NC-ND 4.0.}}
  \label{SEIR:fig:age-contacts}
\end{figure}

@@

MCMC using a Normal proposal distribution can be inefficient.
In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
HMC simulates a physical system travelling through the posterior space with momentum.
The momentum increases in the direction the posterior density increases, using the gradient.
\todo{add something about leapfrog steps}
This means that the sampler tends to propose moves with higher posterior density.
The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.

>>>>

The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocite{beckerDependent,brookmeyerMethod}.

@@

% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
% A super-spreading event is an event where many more people than average are infected.
% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
% %The negative binomial tends to Poisson as $k\to\infty$.
% Generally, $k<1$ is considered as significant levels of overdispersion.
% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocite{quiltyReconstructing,quiltyUnderstanding}.
% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling}{andersonSpread}.
This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The system can now be written as follows:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocite{birrellBayesian,jacksonEffects}.
Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12. A clear diagonal, showing associative mixing is visible, and children have more contacts. Parallel diagonals off the main diagonal show mixing between parents and their children. Figure adapted from \textcite{klepacContacts}, licensed under \href{https://creativecommons.org/licenses/by-nc-nd/4.0/}{CC BY-NC-ND 4.0.}}
  \label{SEIR:fig:age-contacts}
\end{figure}

@@

MCMC using a Normal proposal distribution can be inefficient.
In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
\emph{Hamiltonian Monte Carlo} (HMC)~\autocite{nealMCMC,nealImproved,duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
HMC simulates a physical system travelling through the posterior space with momentum.
The momentum increases in the direction the posterior density increases, using the gradient.
\todo{add something about leapfrog steps}
This means that the sampler tends to propose moves with higher posterior density.
The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.

### 2024-02-01 introduction.tex
preferable citation sorting

\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
\Textcite{brookmeyerMethod} developed it in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome).
HIV is transmitted sexually or through blood and causes the disease AIDS.
It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

>>>>

\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
\Textcite{brookmeyerMethod} developed it in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome).
HIV is transmitted sexually or through blood and causes the disease AIDS.
It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

### 2024-02-01 transmission.tex
preferable citation sorting

During November, incidence starts rising, although starting at different times in different regions.
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocites{walkerTracking}{daviesEstimated}{lythgoeLineage}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
This follows directly from the ODEs if $\vec{s}$ is constant. 
The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
\begin{align}
    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
\end{align}

>>>>

During November, incidence starts rising, although starting at different times in different regions.
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocite{chenRole,vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
This follows directly from the ODEs if $\vec{s}$ is constant. 
The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
\begin{align}
    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
\end{align}
